Reordering Elements in List CRDTs

### 2020-02-20 paper/list-move.tex
Move figure, add related work

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003809.10010172</concept_id>
       <concept_desc>Theory of computation~Distributed algorithms</concept_desc>
       <concept_significance>500</concept_significance>
   </concept>
   <concept>
       <concept_id>10003120.10003130.10003131.10003570</concept_id>
       <concept_desc>Human-centered computing~Computer supported cooperative work</concept_desc>
       <concept_significance>300</concept_significance>
   </concept>
   <concept>
       <concept_id>10002951.10003227.10003233</concept_id>
       <concept_desc>Information systems~Collaborative and social computing systems and tools</concept_desc>
       <concept_significance>300</concept_significance>
   </concept>
   <concept>
       <concept_id>10010520.10010521.10010537.10010540</concept_id>
       <concept_desc>Computer systems organization~Peer-to-peer architectures</concept_desc>
       <concept_significance>300</concept_significance>
   </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Distributed algorithms}
\ccsdesc[300]{Human-centered computing~Computer supported cooperative work}
\ccsdesc[300]{Information systems~Collaborative and social computing systems and tools}
\ccsdesc[300]{Computer systems organization~Peer-to-peer architectures}

@@

    \vspace{2em}

@@

The algorithm of Section~\ref{sec:algorithm} allows a single list element to be moved at a time.
In some applications, a further challenge arises: we may want to move not just one element, but move a whole consecutive range of elements in one operation.
For example, in a text document, a user may want to move a section of text (e.g.\ a paragraph, or the text of a bullet point) to a different position in the text.

We might try to move a range of elements by moving each element individually.
However, this approach has a problem, which is illustrated in Figure~\ref{fig:range-bad}.
In this example, replica $B$ wants to swap the order of two bullet points in the text of a shopping list by moving the character range ``$\bullet$ Milk'' (including the bullet point character $\bullet$), while concurrently replica $A$ changes the text ``Milk'' to ``Soy milk''.
While the move is successful, the character insertions and deletion of $A$'s edit remain attached to the old position of ``Milk'' (at the end of the document, after ``Bacon$\backslash$n'').
This is the case because CRDTs for lists and text perform insertions and deletions at particular position IDs, which remain unchanged even as elements are moved to new position IDs.

@@

>>>>

\begin{CCSXML}
<ccs2012>
    <concept>
        <concept_id>10003752.10003809.10010172</concept_id>
        <concept_desc>Theory of computation~Distributed algorithms</concept_desc>
        <concept_significance>500</concept_significance>
    </concept>
    <concept>
        <concept_id>10011007.10010940.10010992.10010993.10010996</concept_id>
        <concept_desc>Software and its engineering~Consistency</concept_desc>
        <concept_significance>500</concept_significance>
    </concept>
    <concept>
        <concept_id>10003120.10003130.10003131.10003570</concept_id>
        <concept_desc>Human-centered computing~Computer supported cooperative work</concept_desc>
        <concept_significance>300</concept_significance>
    </concept>
    <concept>
        <concept_id>10002951.10003227.10003233</concept_id>
        <concept_desc>Information systems~Collaborative and social computing systems and tools</concept_desc>
        <concept_significance>300</concept_significance>
    </concept>
    <concept>
        <concept_id>10010520.10010521.10010537.10010540</concept_id>
        <concept_desc>Computer systems organization~Peer-to-peer architectures</concept_desc>
        <concept_significance>300</concept_significance>
    </concept>
</ccs2012>
\end{CCSXML}

\begin{CCSXML}
<ccs2012>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Distributed algorithms}
\ccsdesc[500]{Software and its engineering~Consistency}
\ccsdesc[300]{Human-centered computing~Computer supported cooperative work}
\ccsdesc[300]{Information systems~Collaborative and social computing systems and tools}
\ccsdesc[300]{Computer systems organization~Peer-to-peer architectures}

@@

@@

The algorithm of Section~\ref{sec:algorithm} allows a single list element to be moved at a time.
In some applications, a further challenge arises: we may want to move not just one element, but move an entire consecutive range of elements in one operation.
For example, in a text document, a user may want to move a section of text (e.g.\ a paragraph, or the text of a bullet point) to a different position in the text~\cite{Lord:2019bo}.

We might try to move a range of elements by moving each element individually.
However, this approach has a problem, which is illustrated in Figure~\ref{fig:range-bad}.
In this example, replica $B$ wants to swap the order of two bullet points in the text of a shopping list by moving the character range ``$\bullet$ Milk'' (including the bullet point character $\bullet$), while concurrently replica $A$ changes the text ``Milk'' to ``Soy milk''.
While the move is successful, the character insertions and deletion of $A$'s edit remain attached to the old position of ``Milk'' (near the end of the document, after ``Bacon$\backslash$n'').
This is the case because CRDTs for lists and text perform insertions and deletions at particular position IDs, which remain unchanged even as elements are moved to new position IDs.

@@

% bounds of range are given as positions (so that moving a range containing a bound does not move
% the bound somewhere else), but insertions/deletions reference element identity

% Inserted item may be relocated if its insertion position is within a range that was moved. This
% implies the insertion position cannot be given as a position ID in the operation, since a new
% position ID would have to be generated in the effector. However, if insertion contains only
% identity timestamp, as in RGA, then it’s fine?

% But then if a range’s bounds are specified as position IDs, and an insertion contains only an
% identity timestamp, where do the position IDs come from?

% Maybe an insert op should contain two timestamps — one for identity, one for position?

\section{Related work}

Our previous unpublished work \cite{ExtendedVersion} discusses a single-element move operation for lists in the context of a larger specification framework for CRDTs.
However, this work does not provide an efficient implementation of this operation.

Apart from this work, to our knowledge Ahmed-Nacer et al.~\cite{AhmedNacer:2013eq} provide the only published list CRDT that includes a single-element move operation.
However, it is designed to duplicate an element when moved concurrently~-- semantics we consider undesirable.
Moreover, we believe the algorithm to be more complicated than necessary.

All other list CRDTs~\cite{Oster:2006wj,Preguica:2009fz,Roh:2011dw,Grishchenko:2014eh,Attiya:2016kh,Weiss:2009ht,Weiss:2010hx,Nedelec:2013ky,Nedelec:2016eo} and Operational Transformation algorithms \cite{Ellis:1989ue,Nichols:1995fd,Ressel:1996wx,Sun:1998vf,Oster:2006tr} provide only insertion and deletion.

Lord~\cite{Lord:2019bo} discusses the problem of moving ranges of elements, but does not present a solution.
